Applying the optimized ensemble histogram method to the square-well liquid

### 2015-01-30 papers/square-well-liquid/histogram-paper.tex
various cleanups in TMMC method + some actual text describing it Tested on bennet. Took 12:13.97.

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/periodic-ww130-ff30-N20-wang_landau-transitions}\\
  \includegraphics[width=\columnwidth]{figs/periodic-ww130-ff30-N20-tmmc-transitions}\\
  \includegraphics[width=\columnwidth]{figs/periodic-ww130-ff30-N20-wang_landau-tmmc-compare-transitions}
  \caption{A plot showing the probability of energy transitions (with
    no weighting) from given energy states.\label{fig:transitions}}
\end{figure}

@@

\subsubsection{Sampling the transition matrix (more) directly}

If our goal (in initialization) is to sample the transition matrix
(from which we compute the optimal weights), we could use a much more
direct approach.  When computing a histogram, we are forced to perform
a complete random walk.  However, the transition matrix can be
computed at a given energy \emph{without sampling any other energies}!

We can start at a given energy, sample the transition matrix at that
energy to our satisfaction, i.e. until we have observed transitions in
both directions a statistically significant number of times.  Then we
would proceed to the next energy and sample it, and so forth.  There
is a weakness in that we may not be able to sample all states at a
given energy without going to other energies.  But we could repeat
this procedure many times to accomplish this.  In a way, this
algorithm is reminiscent of Wang-Landau, which ``forces'' the
simulation to sample a whole slew of different energies, but in our
case we would be able to put strict bounds on how long it spends
sampling each energy, and could reasonably keep the code from wasting
huge amounts of time sampling the high-entropy states.

>>>>

\begin{figure*}
  \includegraphics[width=0.33\textwidth]{figs/periodic-ww130-ff30-N20-wang_landau-transitions}\hfill%
\includegraphics[width=0.33\textwidth]{figs/periodic-ww130-ff30-N20-tmmc-transitions}\hfill%
\includegraphics[width=0.33\textwidth]{figs/periodic-ww130-ff30-N20-wang_landau-tmmc-compare-transitions}
  \caption{A plot showing the probability of energy transitions (with
    no weighting) from given energy states.\label{fig:transitions}}
\end{figure*}

@@

\subsubsection{Initializing using the transition matrix directly}

An alternative initialization algorithm involves using the transition
matrix directly during initialization.  This process was proposed by
Swendsen \emph{et al.} as a modification for their initialization
procedure~\cite{swendsen1999transition}.  In this case, the transition
matrix is continually updated, thus modifying the relative weighting
of different energies as the simulation proceeds.  This strictly
speaking no longer satisfies detailed balance, but as the simulation
proceeds this violation will quickly become increasingly small.

This raises the question of how to use the transition matrix to
directly determine transition probabilities, since the self-consistent
eigenvalue solution is unsuitable.  The key is to recognize that we do
not require weights for each energy but only \emph{differences} of
weights for on pair of energies at a time.  This is a much smaller
problem.

Swendsen \emph{et al.} find for a flat distribution, the acceptance
rate for a given transition is equal to the ratio of that element in
the transition matrix to the element corresponding to the reverse
transition~\cite{swendsen1999transition}.  Naturally, for a different
distribution, we can scale the acceptance ratios according to the
ratio of occupations desired.  We modify this algorithm to always
accept transitions to a lower energy, thus employing a broad histogram
method that does not much sample energies above the maximum-entropy
energy.

Swendsen only uses this approach after a slightly hokey two-stage
initialization.  It seems to work pretty well, and asymptotically it
does have detailed balance.
